\bf Title here

### 2020-07-29 paper/paper.tex
modify based on meeting 19 and polish section 3 and 4 (except 4.1)

%% load any required packages here
\usepackage{amssymb, amsmath, mathtools, dsfont} \usepackage[ruled,vlined, linesnumbered]{algorithm2e}

@@

The paper is organised as follows. Section \ref{DFO} gives a general
literature review of optimisation, specifically derivative free
optimisation. Section \ref{vis-diag} presents the new idea of
constructing a systematic visual framework that diagnoses the components
of an optimisation procedure (parameters, searching path, etc). The rest
of the paper serves as a comprehensive example of using the visual
diagnostics on one particular problem: \emph{projection pursuit guided
tour}. Some background knowledge of projection pursuit guided tour is
provided in Section \ref{tour}. Section \ref{apply} applies the concepts
proposed in section \ref{vis-diag} in the tour problem and sets up the
data. The last section, Section \ref{plots}, presents the visual
diagnostic plots and explains how they can help to understand different
aspects of the optimisation in tour.

@@

From Section \ref{tour}, we presents a comprehensive case study on how
to use the visual diagnostics to explore the optimisation in a specific
problem: projection pursuit guided tour. Section \ref{tour} aims to
provide non-experts with an overview of the problem content and the
existing optimisation procedures used in projection pursuit guided tour.
For those who are already familiar with the techniques, feel free to
skip this section.

The optimisation problem we're interested in is in the context of
projection pursuit. Coined by \citet{friedman1974projection}, projection
pursuit is a method that detects the interesting structure
(i.e.~clustering, outliers and skewness) of multivariate data via
projecting it in lower dimensions. Let \(\mathbf{X}_{n \times p}\) be a
data matrix, an n-d projection can be seen as a linear transformation
\(T: \mathbb{R}^p \mapsto \mathbb{R}^d\) defined by
\(\mathbf{P} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{P}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis. Define
\(f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}\) be an index function
that maps the projection basis \(\mathbf{A}\) onto a scalar value \(I\),
this function is commonly known as the projection pursuit index (PPI)
function, or the index function and can be used to measure the
``interestingness'' of the projection. A number of index functions have
been proposed in the literature including Legendre index
\citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial}, natural Hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}.

As \citet{friedman1974projection} noted ``\ldots{}, the technique use
for maximising the projection index strongly influences both the
statistical and the computational aspects of the procedure.'' A suitable
optimisation procedure is needed to find the projection angle that
maximises the PPI and the quality of the optimisation largely affect the
interesting projections one could possibly observe.

Projection pursuit is usually used in conjunction with a tour method
called \emph{guided tour}. Tour explores the multivariate data
\emph{interactively} via playing a series of projections, that form a
\emph{tour path} and guided tour uses the path that is geodesically the
shortest. Details of the mathematical construction of a tour path can be
found in \citet{buja2005computational}. {[}Maybe add the sketch for
illustrating tour path.{]} Guided tour, along with other types of tour,
has been implemented in the \emph{tourr} package in R, available on the
Comprehensive R Archive Network at
\url{https://cran.r-project.org/web/packages/tourr/}
\citep{wickham2011tourrpackage}.

\begin{figure}
\includegraphics[width=1\linewidth,height=0.6\textheight]{/Users/hzha400/Documents/3.PhD/research/paper-tour-vis/figures/tour_path_increasing} \caption{\label{tour-path}An illustration of the tour path}\label{fig:tour-path}
\end{figure}

\hypertarget{optimisation-problem-and-existing-algorithms}{%
\subsection{Optimisation problem and existing
algorithms}\label{optimisation-problem-and-existing-algorithms}}

Now we begin to formulate the optimisation problem. Given a randomly
generated starting basis \(\mathbf{A}^0\), projection pursuit finds the
final projection basis
\(\mathbf{A}^J = [\mathbf{a}_1, \cdots, \mathbf{a}_d]\), where
\(\mathbf{a}_i\) is a \(p \times 1\) vector, satisfies the following
optimisation problem:

@@

via an iterative search of target basis:
\(\{\mathbf{A}^0, \mathbf{A}^1, \mathbf{A}^2, \cdots \cdots \mathbf{A}^J\}\).

There are three existing methods for optimising PPI function and we
review them below.

\citet{posse1995projection} proposed a stochastic direct search method,
a random search algorithm. Given a current basis
\(\mathbf{A}_{\text{cur}}\), a candidate basis \(\mathbf{A}_{t}\) is
sampled in the neighbourhood defined by the radius of the p-dimensional
sphere, \(\alpha\) of the current basis \(\mathbf{A}_{t}\) by
\(\mathbf{A}_{t} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}\).
If the candidate basis has a higher index value than the current basis,
it is outputted as the target basis \(\mathbf{A}_{t}\) along with other
metadata and the current iteration stops. If no basis is found to have
higher index value after the maximum number of tries \(l_{\max}\), the
algorithm stops with nothing outputted. \(c\) is the halfing parameter
and when \(c > 30\), the neighbourhood parameter \(\alpha\) in the next
iteration will be reduced by half. The algorithm for a random search is
summarised in Algorithm \ref{random-search}.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The target basis: $\mathbf{A}_{t+1}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{l}$ is orthonormal\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    \eIf{$I_{l} > I_{\text{cur}}$}{
      \KwRet{$\mathbf{A}_{l}$} \;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

\citet{cook1995grand} explained the use of a gradient ascent
optimisation with the assumption that the index function is continuous
and differentiable. Since some indices could be non-differentiable, the
computation of derivative is replaced by a pseudo-derivative of
evaluating five randomly generated directions in a tiny nearby
neighbourhood. Taking a step on the straight derivative direction has
been modified to maximise the projection pursuit index along the
geodesic direction. See Algorithm \ref{search-geodesic} for a
summarisation of the steps.

@@

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated} is a
non-derivative procedure based on a non-increasing cooling scheme
\(T(t)\). Given an initial \(T_0\), the temperature at iteration \(t\)
is defined as \(T(t) = \frac{T_0}{log(t + 1)}\). The simulated annealing
algorithm works as follows. Given a neighbourhood parameter \(\alpha\)
and a randomly generated orthonormal basis \(A_{rand}\), a candidate
basis is constructed as
\(\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{rand}\).
If the index value of the candidate basis is larger than the one of the
current basis, the candidate basis becomes the target basis. If it is
smaller, the candidate is accepted with probability
\(P= \min\left\{\exp\left[\frac{I_{t+1} - I_{t}}{T(t)}\right],1\right\}\)
where \(I_{t}\) and \(I_{t+1}\) are the index value of the current and
candidate basis respectively. The algorithm can be summarised as in
Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Set $l = 0$\;
    Generate $\mathbf{A}_{l} = (1 - \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{l}$ is orthonormal \;
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

Below listed several features characterise the optimisation procedures
needed in projection pursuit

\begin{itemize}
\item
  \emph{Being able to handle non-differentiable PPI function}: The PPI
  function could be non-differentiable, thus derivative free methods are
  preferred.
\item
  \emph{Being able to optimise with constraints}: The constraint comes
  from projection matrix being an orthonormal matrix.
\item
  \emph{Being able to find both local and global maximum}: Although the
  primary interest is to find the global maximum, local maximum could
  also reveal structures in the data that are of our interest.
\end{itemize}

@@

Random search methods has a black-box mechanism and focuses solely on
finding the global maximum point, while the projection pursuit problem
we have aims at \emph{exploring} the data and thus is interested in how
the algorithm finds its maximum. This motivates us \textbf{to develop a
visual diagnostic system for exploring the optimisation searching path}.

@@

Ggplot requires a data frame that contains all the variables to plot and
a \emph{global object} is constructed as the data frame supplied to the
visual diagnostic plots to better suit the characters of iterative
optimisation algorithms. Given an optimisation algorithm, two primary
variables of interest are the \emph{decision variable: \(\mathbf{A}\)}
and the \emph{value of the objective function: \(I\)}, which is the
projection basis and index value repsectively in projection pursuit.
\emph{State: \(S\)} is another variable of great importance since it
labels \(\mathbf{A}\) and \(I\) by the stage they are in the whole
search-and-interpolate process.

\emph{Iterators} indexes the data collected and prescribes the the
natural ordering of each observation in different levels. \emph{\(t\)}
is a unique identifier that has an increment of one for each observation
and \emph{\(j\)} is the counter for each search-and-interpolate round.
{[}add a bit more for \emph{\(l\)} after decide whether to combine \(l\)
and \(k\). {]}

There are other parameters that are of our interest and we denote them
as \emph{\(V_{p}\)}. In projection pursuit, this includes
\(V_1 = \text{method}\), denoting the name of the searching method used
and \(V_2 = \text{alpha}\), which is a neighbourhood parameter that
controls the size in sampling candidate bases. A sketch of the global
object for projection pursuit guided tour is presented in Figure
\ref{fig:glb-obj}. The full data structure can thus be shown as in
Equation \ref{eq:data-structure}.

\begin{equation}
\left[
\begin{array}{c|ccc|ccc|ccc}
t & \mathbf{A} & I & S & j &  l  & k & V_{1}& \cdots & V_{p}\\
\hline
1 & \mathbf{A}_1 & I_1 & S_1 & 1 & 1 & \ast & V_{1, 1} & \cdots & V_{p, 1}\\
\hline
2 & \mathbf{A}_2 & I_2 & S_2 & 2 & 1 & \ast & V_{1, 2} & \cdots & V_{p, 2}\\
3 & \mathbf{A}_3 & I_3 & S_3 & 2 & 2 & \ast & V_{1, 3} & \cdots & V_{p, 3}\\
\vdots & \vdots &\vdots &\vdots & \vdots &\vdots & \vdots &\vdots &\cdots &\vdots\\
\vdots & \vdots & \vdots &\vdots & 2 & l_1 & \ast & \vdots &  \cdots & \vdots\\
\hline
\vdots &\vdots & \vdots &\vdots & 2 & \ast & 1& \vdots &  \cdots & \vdots\\
\vdots &\vdots &\vdots &\vdots & 2 & \ast & 2& \vdots& \cdots & \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
\vdots &\vdots &\vdots &\vdots & 2 & \ast & k_1&\vdots & \cdots & \vdots\\
\hline
\vdots &\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
T & \mathbf{A}_T & I_T &S_T  & J &  l_J & \ast & V_{1, T}& \cdots & V_{p, T}\\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}

@@

We first simulate some random variables of size 1000 with different
structures. \texttt{x1}, \texttt{x8}, \texttt{x9} and \texttt{x10} are
simulated from normal distribution with zero mean and variance of one as
in equation \ref{eq:sim-norm}. When using projection pursuit to explore
the data structure based on its departure from normality, the entry in
the projection basis for these variables should be close to zero in
theory. \texttt{x2} to \texttt{x7} are mixture of normal distribution
with different weights and locations. Equation \ref{eq:sim-x2} to
\ref{eq:sim-x7} show the distribution of each variables and Figure
\ref{origin-data} shows the histogram of each variable except
\texttt{x3}. All the variables are then scaled to ensure the mixture has
variance of one.

@@

We form our first dataset using variable \texttt{x1}, \texttt{x2},
\texttt{x8}, \texttt{x9} and \texttt{x10} and supply it to one of the
above algorithm, say \texttt{search\_geodesic} in the \texttt{tourr}
package. When the optimisation ends, the global object will be stored
and printed (it can be turned off by supplying argument
\texttt{print\ =\ FALSE}). Additional messages during the optimisation
can be displayed by \texttt{verbose\ =\ TRUE}. Below shows the first
five rows of the global object. The global object meets the definition
of tidy data introduced in \citep{wickham2014tidy} where each
observation forms a row, each variable forms a column and each type of
observational unit forms a table. The adoption of tidy data makes it
easier for data wrangling and visualisation, which powers the visual
diagnostics system introduced in this paper.

@@

\begin{verbatim}
## # A tibble: 5 x 8
##   basis             index_val tries info           loop method       alpha    id
##   <list>                <dbl> <dbl> <chr>         <dbl> <chr>        <dbl> <int>
## 1 <dbl[,1] [5 x 1]>     0.749     1 start            NA <NA>           0.5     1
## 2 <dbl[,1] [5 x 1]>     0.730     1 random_search     1 search_bett~   0.5     2
## 3 <dbl[,1] [5 x 1]>     0.743     1 random_search     2 search_bett~   0.5     3
## 4 <dbl[,1] [5 x 1]>     0.736     1 random_search     3 search_bett~   0.5     4
## 5 <dbl[,1] [5 x 1]>     0.747     1 random_search     4 search_bett~   0.5     5
\end{verbatim}

\hypertarget{visual-diagnostics}{%
\section{Visual diagnostics}\label{visual-diagnostics}}

@@

  \begin{itemize}
  \tightlist
  \item
    data: group by \(j\) and filter the observations in the group that
    have count greater than \texttt{cutoff\ =\ 15}.
  \item
    x: \(j\) is mapped to the x-axis
  \item
    y: the statistical transformed index value:
    \(Q_q(I_{T} \cdot \mathds{1} (S = \text{search}))\) is mapped to the
    y-axis where \(q\) takes one of \(0, 25, 50, 75, 100\) and
    \(Q_q(x)\) finds the qth-quantile for the vector \(x\).
    \(\mathds{1}(.)\) is the identity operator.
  \end{itemize}
\item
  Layer 2: point geom

@@

\hypertarget{implementation-ferrn-package}{%
\section{Implementation: Ferrn
package}\label{implementation-ferrn-package}}

>>>>

%% load any required packages here
\usepackage{amssymb, amsmath, mathtools, dsfont, bbm} \usepackage[ruled,vlined, linesnumbered]{algorithm2e}

@@

The remainder of the paper is organised as follows. Section \ref{DFO}
gives a general overview of optimisation methods, specifically
derivative free optimisation. Section \ref{tour} introduces the
optimisation problem in a specific problem: projection pursuit guided
tour and existing derivative free algorithms for solving it. Section
\ref{vis-diag} and \ref{vis-diag-plots} together present the new
construction of a visual diagnostic framework that unveil the blackbox
of optimisation. Section \ref{vis-diag} focues on motivational ideas and
data structure and Section \ref{vis-diag-plots} provides definitions of
different diagnostic plots using grammar of graphic and examples.
Finally, \ref{implementation} mentions the R package: ferrn that
implements the visual diagnostics described above.

@@

The optimisation problem we're interested in is in the context of
projection pursuit guided tour. Coined by
\citet{friedman1974projection}, projection pursuit detects interesting
structures (i.e.~clustering, outliers and skewness) in multivariate data
via projecting it in lower dimensions. Let \(\mathbf{X}_{n \times p}\)
be the data matrix, an n-d projection can be seen as a linear
transformation \(T: \mathbb{R}^p \mapsto \mathbb{R}^d\) defined by
\(\mathbf{P} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{P}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis. Define
\(f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}\) be an index function
that maps the projection basis \(\mathbf{A}\) onto a scalar index value
\(I\), this function is commonly known as the projection pursuit index
(PPI) function, or the index function and is used to measure the
``interestingness'' of the projection. A number of index functions have
been proposed in the literature to detect different data structures,
including Legendre index \citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial}, natural Hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}.

In their initial paper, \citet{friedman1974projection} noted that
``\ldots{}, the technique use for maximising the projection index
strongly influences both the statistical and the computational aspects
of the procedure.'' Hence, effective optimisation algorithms are
necessary for projection pursuit to find bases that give interesting
projections. While we leave the formal construction the optimisation
problem and existing algorithms in section \ref{tour-optim}, we want to
emphasizes here that given a random starting (current) basis, projection
pursuit repetitively searches for candidate bases nearby and outputs a
target basis once it finds one with higher index value than the current
basis. The target basis becomes the current basis in the next iteration
and the searching continues until no better basis can be found or the
termination criteria are reached.

A tour method, guided tour is usually used in conjunction with
projection pursuit to interactively explore multivariate data. Tour
plays an animation on the projected data which is constructed using the
bases outputted by projection pursuit and their interpolations. Guided
tour constructs the interpolation using the shortest geodesic path
between two outputted bases and we refer readers to
\citet{buja2005computational} for details of the mathematical
construction of the interpolation basis Figure \ref{tour-path} shows a
sketch of the tour path consisting of the blue frames produced by
projection pursuit and the white frames, which interpolates two blue
frames. The tour method has been implemented in the \emph{tourr} package
in R, available on the Comprehensive R Archive Network at
\url{https://cran.r-project.org/web/packages/tourr/}
\citep{wickham2011tourrpackage}.

\begin{figure}
\includegraphics[width=1\linewidth,height=0.6\textheight]{/Users/hzha400/Documents/3.PhD/research/paper-tour-vis/figures/tour_path_keynote/tour_path_keynote.001} \caption{\label{tour-path}An illustration of the tour path}\label{fig:tour-path}
\end{figure}

\hypertarget{tour-optim}{%
\subsection{Optimisation problem and existing
algorithms}\label{tour-optim}}

Now we begin to formulate the optimisation problem. Given a randomly
generated starting basis \(\mathbf{A}_1\), projection pursuit finds the
final projection basis
\(\mathbf{A}_T = [\mathbf{a}_1, \cdots, \mathbf{a}_d]\), where
\(\mathbf{a}_i \in \mathbb{R}^{p}\), satisfies the following
optimisation problem:

@@

There are three existing algorithms.

\citet{posse1995projection} proposed a random search algorithm that
samples a candidate basis \(\mathbf{A}_{l}\) in the neighbourhood of the
current basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}\),
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{rand}\) is a randomly generated matrix with the same
dimension as \(\mathbf{A}_{\text{cur}}\). The optimiser keeps sampling
bases near the current basis until it finds one with higher index value
than the current basis and then outputs it for guided tour to construct
the interpolation path. A new round of search continues to find a better
basis after the interpolation finishes. The halfing parameter \(c\) with
default value of 30 is designed to adjust the searching neighbourhood
\(\alpha\). When the search needs to sample more than \(c\) number of
basis to find an outputted basis, the neighbourhood parameter \(\alpha\)
will be reduced by half in the next iteration. If the optimiser can't
find a better basis within the maximum number of tries \(l_{\max}\), the
algorithm stops. The algorithm is summarised in Algorithm
\ref{random-search} for one iteration.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The target basis: $\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$ and $c = 0$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{l}$ is orthonormal\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    \eIf{$I_{l} > I_{\text{cur}}$}{
      \KwRet{$\mathbf{A}_{l}$} \;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

\citet{cook1995grand} explained the use of a gradient ascent
optimisation with an assumption that the index function is continuous
and differentiable. Since some indices could be non-differentiable, the
derivative is replaced by a pseudo-derivative of evaluating five
randomly generated directions in a tiny nearby neighbourhood. Once the
best direction is determined, the optimiser finds the best projection
\(\mathbf{A}_{**}\) on the geodesic between the current basis
\(\mathbf{A}_{\text{cur}}\) and the best directional basis
\(\mathbf{A}_{*}\). If the percentage change in the index value between
\(\mathbf{A}_{**}\) and \(\mathbf{A}_{\text{cur}}\) is greater than a
threshold value, \(\mathbf{A}_{**}\) is outputted for the current
iteration to construct the interpolation or the algorithm repeats the
above steps until \(l_{\max}\) is reached and the searching terminates.
Algorithm \ref{search-geodesic} summarise the steps in geodesic search.

@@

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated} is a
non-derivative procedure based on a non-increasing cooling scheme
\(T(l)\). Given an initial \(T_0\), the temperature at iteration \(l\)
is defined as \(T(l) = \frac{T_0}{\log(l + 1)}\). The simulated
annealing algorithm works as follows. Given a neighbourhood parameter
\(\alpha\) and a randomly generated orthonormal basis \(A_{rand}\), a
candidate basis is sampled by
\(\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{rand}\).
If the index value of the candidate basis is larger than the one of the
current basis, the candidate basis is outputted. While if it is smaller,
the candidate has another chance of being accepted with probability
\(P= \min\left\{\exp\left[\frac{I_{cur} - I_{l}}{T(l)}\right],1\right\}\)
where \(I_{\cdot}\) denotes the index value of a given basis. The
algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$; Initial termperature: $T_0$}
    \Output{The global object; The target basis: $\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = (1 - \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{l}$ is orthonormal \;
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

Notice that the three algorithms above constitute one search. The
outputted basis is then sent to construct the interpolation and one such
search-and-interpolate step is one round.

@@

Random search methods has a black-box mechanism and focuses solely on
finding the global maximum point, while the projection pursuit explores
the data with an aim to discover the potential underlying structure and
thus is interested in how the algorithm finds its maximum. This
motivates us \textbf{to develop a visual diagnostic system for exploring
the optimisation searching path}.

@@

Tidy data, as introduced in \citep{wickham2014tidy}, has three criteria:
1) each observation forms a row, 2) each variable forms a column and 3)
each type of observational unit forms a table. The adoption of tidy data
makes it easier for data wrangling and visualisation. For our visual
diagnostics, global object is a such data construction with key
components in optimisation algorithms and is designed for making
diagnostic plots easier. In the optimisation algorithms for projection
pursuit, two key elements are the projection basis: \(\mathbf{A}\) and
the index value: \(I\). State: \(S\) labels the observation with
detailed stage of searching or interpolation.

Multiple iterators are needed to index the data collected at different
levels. \(t\) is a unique identifier that prescribes the natural
ordering of each observation; \(j\) is the counter for each
search-and-interpolate round, which remains the same for within one
round and has an increment of one once a new round starts. \(l\) is the
counter for each search/interpolation allowing us to know how many basis
has the algorithm searched before finding one to output.

There are other parameters that are of our interest and we denote them
as \emph{\(V_{p}\)}. In projection pursuit, this includes
\(V_1 = \text{method}\), which tags the name of the algorithm used and
\(V_2 = \text{alpha}\), which is the neighbourhood parameter that
controls the size in sampling candidate bases. The data structure can
thus be shown as in Equation \ref{eq:data-structure}.

\begin{equation}
\left[
\begin{array}{c|ccc|cc|cc}
t & \mathbf{A} & I & S & j &  l  & V_{1} & V_{2}\\
\hline
1 & \mathbf{A}_1 & I_1 & S_1 & 1 & 1 & V_{11} & V_{21}\\
\hline
2 & \mathbf{A}_2 & I_2 & S_2 & 2 & 1  & V_{12}  & V_{22}\\
3 & \mathbf{A}_3 & I_3 & S_3 & 2 & 2  & V_{13}  & V_{23}\\
\vdots & \vdots &\vdots &\vdots  &\vdots & \vdots &\vdots  &\vdots\\
\vdots & \vdots & \vdots &\vdots & 2 & l_2 & \vdots  & \vdots\\
\hline
\vdots &\vdots & \vdots &\vdots & 2  & 1& \vdots & \vdots\\
\vdots &\vdots &\vdots &\vdots & 2 & 2& \vdots &  \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots & \vdots  &\vdots \\
\vdots &\vdots &\vdots &\vdots & 2 & k_2 &\vdots  & \vdots\\
\hline
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
\hline
\vdots & \vdots & \vdots &\vdots  & J &  1 & \vdots & \vdots \\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
T & \mathbf{A}_T & I_T &S_T  & J &  l_{J} & V_{1T}& V_{2T}\\
\hline
\vdots &\vdots & \vdots &\vdots & J  & 1& \vdots & \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots & \vdots  &\vdots \\
\vdots &\vdots &\vdots &\vdots & J & k_J &\vdots  & \vdots\\
\hline
\vdots& \vdots & \vdots & \vdots & J+1 & 1 & \vdots& \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
T^\prime & \mathbf{A}_{T^\prime} & I_{T^\prime} &S_{T^\prime}  & J+1 &  l_{J+1} & V_{1T^\prime}& V_{2T^\prime}\\
\end{array}
\right]
= 
\left[
\begin{array}{c}
\text{column name} \\
\hline
\text{search (start basis)} \\
\hline
\text{search} \\
\text{search} \\
\vdots \\
\text{search (outputed basis)} \\
\hline
\text{interpolate} \\
\text{interpolate} \\
\vdots \\
\text{interpolate} \\
\hline
\vdots \\
\hline
\text{search} \\
\vdots \\
\text{search (final basis)} \\
\hline
\text{interpolate} \\
\vdots \\
\text{interpolate} \\
\hline
\text{search (no output)} \\
\vdots \\
\text{search (no output)} \\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}

where \(T^{\prime} = T + k_{J}+ l_{J+1}\). Note that we deliberately
denote the last round of search as \(j = J+1\) and in that round there
is no output/interpolation basis and the algorithm terminates. This
notation allows us to denote the last complete search-and-interpolate
round as round \(J\) and hence the final basis is \(A_T\) and highest
index value found is \(I_T\).

{[}outside the paper: I find the notation of current/target basis is
confusing because the target basis in round \(j\) becomes the current
basis in round \(j+1\). Also, when we start to have polish, the target
basis may not be the current basis in the next round\ldots{} The place
where current/target is most appropriate is probably when describing the
interpolation where the first one is always the current basis and the
last is always the target basis. I think it is better to leave this
language in the code{]}

A sketch of the global object for projection pursuit guided tour is
presented in Figure \ref{fig:glb-obj}.

{[}I feel this sketch was initially useful but now since we have the
data matrix and the printed output of the global object, it doesn't any
additional information. I'm still keeping it here but we may need to
remove it if it's not useful :({]}

@@

We simulate some random variables of size 1000 with different
structures. \texttt{x1}, \texttt{x8}, \texttt{x9} and \texttt{x10} are
simulated from normal distribution with zero mean and variance of one as
in equation \ref{eq:sim-norm}. When using projection pursuit to explore
the data structure based on its departure from normality, the entry in
the projection basis for these variables should be close to zero in
theory. \texttt{x2} to \texttt{x7} are mixture of normal distributions
with different weights and locations. Equation \ref{eq:sim-x2} to
\ref{eq:sim-x7} outlines the distribution where each variable is
simulated from and Figure \ref{origin-data} shows the histogram of each
variable except \texttt{x3}. All the variables are then scaled to have
variance of one before running the projection pursuit.

@@

We form our first dataset using variable \texttt{x1}, \texttt{x2},
\texttt{x8}, \texttt{x9} and \texttt{x10} and run the guided tour with
optimiser \texttt{search\_better}. When the optimisation ends, the
global object will be stored and printed (it can be turned off by
supplying argument \texttt{print\ =\ FALSE}). Additional messages during
the optimisation can be displayed by argument \texttt{verbose\ =\ TRUE}.
Below shows the first ten rows of the global object. Notice that the
tibble object allows the list-column \texttt{basis} to be printed out
nicely with the dimension of the projection basis readily available.

{[}I notice that in the implementation the loop index for interpolation
starts from 0 rather than 1. Need more investigation to see if it is
easy to change it to 1 or we will have to change the definition of the
data structure.{]}

@@

\begin{verbatim}
## # A tibble: 10 x 8
##       id basis             index_val info         tries  loop method       alpha
##    <int> <list>                <dbl> <chr>        <dbl> <dbl> <chr>        <dbl>
##  1     1 <dbl[,1] [5 x 1]>     0.749 new_basis        1     1 <NA>           0.5
##  2     2 <dbl[,1] [5 x 1]>     0.730 random_sear~     2     1 search_bett~   0.5
##  3     3 <dbl[,1] [5 x 1]>     0.743 random_sear~     2     2 search_bett~   0.5
##  4     4 <dbl[,1] [5 x 1]>     0.736 random_sear~     2     3 search_bett~   0.5
##  5     5 <dbl[,1] [5 x 1]>     0.747 random_sear~     2     4 search_bett~   0.5
##  6     6 <dbl[,1] [5 x 1]>     0.725 random_sear~     2     5 search_bett~   0.5
##  7     7 <dbl[,1] [5 x 1]>     0.752 new_basis        2     6 search_bett~   0.5
##  8     8 <dbl[,1] [5 x 1]>     0.749 interpolati~     2     0 search_bett~  NA  
##  9     9 <dbl[,1] [5 x 1]>     0.750 interpolati~     2     1 search_bett~  NA  
## 10    10 <dbl[,1] [5 x 1]>     0.750 interpolati~     2     2 search_bett~  NA
\end{verbatim}

\hypertarget{vis-diag-plots}{%
\section{Visual diagnostic plots}\label{vis-diag-plots}}

@@

  \begin{itemize}
  \tightlist
  \item
    data: group by \(j\) and filter the observations in the group that
    have count greater than \texttt{cutoff\ =\ 15}.
  \item
    x: \(j\) is mapped to the x-axis
  \item
    y: the statistical transformed index value: \(Q_q(I^{\prime}_t)\) is
    mapped to the y-axis where \(Q_q(x)\) finds the
    \(0, 25, 50, 75, 100\)th-quantile for the vector \(x\) and
    \(I^{\prime}_t\) is the index value of all the searching bases
    defined in Matrix \ref{eq:data-structure}.
  \end{itemize}
\item
  Layer 2: point geom

@@

\hypertarget{implementation}{%
\section{Implementation: Ferrn package}\label{implementation}}
